---
title: "Differential Expression Analysis of LncRNAs in Skeletal Muscle Following Acute Aerobic Exercise and Acute Resistance Exercise" 
output: pdf_document
author: Kassia Régnier
---

# **Summary**
The following code is the R code used for the differential expression analysis performed in this thesis.   

# Loading package libraries 
```{r, message=FALSE,warning=FALSE}
library(Biobase)
library(BiocManager)
library(dplyr)
library(tidyverse)
library(tibble)
library(GEOquery)
library(DESeq2)
library(ggplot2)
library(biomaRt)
library(pheatmap)
library(RColorBrewer)
```

# Reading in count and phenotype data 
Here, we are reading in the gene expression data (GSE107934) using the GEOquery package.  
```{r, message=FALSE}
pdata_dson <- getGEO("GSE107934")
pdata_dson <- pdata_dson[[1]] #access the expression set 
citation("GEOquery")
```

## Organizing sample information 
Then, we select relevant columns from the pData for our analysis, and create a table.  
```{r,message=FALSE}
pheno_dson <- pData(pdata_dson) %>% 
  dplyr::select(geo_accession,
                `exercise:ch1`,
                `time:ch1`)
```

We rename two of the columns. 
```{r,message=FALSE}
pheno_dson$exercise <- pheno_dson$`exercise:ch1`
pheno_dson$time <- pheno_dson$`time:ch1`
```

Now, we remove the old columns that have been renamed, leaving us with one column for time and one column for exercise.  
```{r,message=FALSE}
pheno_dson_clean = subset(pheno_dson, select = -c(`exercise:ch1`,`time:ch1`))
```

## Designating conditions
We have five conditions to designate: Baseline, 1 hour post-AE, 4 hours post-AE, 1 hour post-RE, and 4 hours post-RE. 
We must create factors after designating conditions, and input the factors into our table of phenotype data. 
We then specify the reference level (Baseline) and include it in our table.  

```{r,message=FALSE}
treatment_dson <- c("Baseline", "AE_1", "AE_4", "RE_1","RE_4")
rep_treat_dson <- c(rep("Baseline",1), rep("AE_1",1), rep("AE_4",1), rep("RE_1",1),rep("RE_4",1),
                    rep("Baseline",1), rep("AE_1",1), rep("AE_4",1), rep("RE_1",1),rep("RE_4",1),
                    rep("Baseline",1), rep("AE_1",1), rep("AE_4",1), rep("RE_1",1),rep("RE_4",1),
                    rep("Baseline",1), rep("AE_1",1), rep("RE_1",1),rep("RE_4",1),
                    rep("Baseline",1), rep("AE_1",1), rep("AE_4",1), rep("RE_1",1),rep("RE_4",1),
                    rep("Baseline",1), rep("AE_1",1), rep("AE_4",1), rep("RE_1",1),rep("RE_4",1)) 

factor_treat <- factor(rep_treat_dson) 
pheno_dson_clean$treatment <- factor_treat
releveld_treat <- relevel(pheno_dson_clean$treatment, ref = "Baseline")
pheno_dson_clean$treatment = releveld_treat 
``` 

# Reading in count data 
Now, we obtain the raw count data. Once read into R, the files must be untar'd before continuing. 
```{r,eval=FALSE, message=FALSE}
getGEOSuppFiles("GSE107934") 
untar("GSE107934/GSE107934_RAW.tar", exdir = "dson") 
```

```{r,message=FALSE}
my_files <- list.files("dson", full.names = TRUE) 
```

Here, we read our files and create a data frame to assess the files. 
```{r,message=FALSE}
count_read <- lapply(my_files, read.table) 
count_df <- as.data.frame(count_read) 
```

## Organizing count data 
Once we have read our files, we organize the count data into a count matrix. 
Then, we organize our count matrix. We rename the row names as Ensembl gene ID's and select only the columns that contain expression data. 
We make sure the row names of our phenotype data and the column names of our count matrix are the same. 
```{r,message=FALSE}
rownames(count_df) <- count_df$V1 
```

```{r,message=FALSE}
new_count_df <- count_df %>%
  dplyr::select(everything()[c(FALSE,TRUE)]) 
```

```{r,message=FALSE, eval=TRUE}
new_count_df <- new_count_df[-c(2,4,6,8,10,
                                12,14,16,18,20,
                                22,24,26,28,30,
                                32,34,36,38,40,
                                42,44,46,48,50,
                                52,54,56,58)] 
cts_dson <- as.matrix(new_count_df) 
colnames(cts_dson) = rownames(pheno_dson_clean) 


```

# Creating the DESEqDataSet 
Following this, we create a DESeqDataSet using the count matrix, pheotype data, and study design (treatment).  
```{r,message=FALSE}
dds_dson <- DESeqDataSetFromMatrix(countData = cts_dson,
                                   colData = pheno_dson_clean,
                                   design = ~treatment)    
```

## Removing genes that are lowly expressed 
We start the DESeqDataSet analysis by pre-filtering the low counts, keeping counts above 1. There are several reasons for this step. By reducing low counts, we decrease the memory size of the DESeqDataSet, and increase the speed of the transformation and testing functions in the DESeq2 package. Also, genes/transcripts having low counts are usually considered artifacts and removing them permits the mean-variance relationship in our data to be more accurately estimated (Law et al., 2016).
```{r,message=FALSE}
keep <- rowSums(counts(dds_dson)) > 1
dds <- dds_dson[keep,]
```

# Differential Expression Analysis 
Here, we perform the differential expression analysis. The function `DESeq()` performs the analysis in 3 steps:
1. estimation of size factors: `estimateSizeFactors`
2. estimation of dispersion: `estimateDispersions`
3. Negative binomial generalized linear model fitting and Wald statistics: `nbinomWaldTest`
```{r,message=FALSE}
dds. <- DESeq(dds)
```

## Obtaining the results names 
We provide `resultsNames()` to obtain the names of the coefficients.  
```{r,message=FALSE}
resultsNames(dds.) 
``` 

Then, we look at the distribution of our normalized count data. 
```{r,echo=TRUE}
boxplot(counts(dds., normalized = TRUE)) 
```

# Exploring and transforming the data 
As RNA-seq data is often heteroskedastic, a variance stabilizing transformation (VST) (Tibshirani 1988; Huber et al. 2003; Anders and Huber 2010) is performed on the count data to obtain count values that are approximately homoskedastic. 
```{r,message=FALSE}
vst <- varianceStabilizingTransformation(dds.) 
```

Boxplot of the VST data. 
```{r,echo=TRUE}
boxplot(assay(vst))
```

## Principal Component Analysis (PCA) 
We can now perform Principal Component Analysis (PCA) (with time and exercise as the groups of interests) to reduce the dimensionality of the data and observe trends, clusters, and outliers in our data. Here, the closer two points are, the more similar they are.   
```{r,echo=TRUE}
plotPCA(vst, intgroup = c("time", "exercise")) 
```

We can modify our PCA plot using `ggplot`. 
```{r,echo=TRUE}
pcaData <- plotPCA(vst, intgroup = c( "time", "exercise"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(x = PC1, y = PC2, color = time, shape = exercise)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA with VST data")
``` 

# Computing results
Now, we compute the results for each coefficient (each condition). 
`results` provides us with a table of results. We can summarize some basic results using `summary`. Here, we provide the summary for only one condition.      
```{r,echo=TRUE}
re.ae.b.1 <- results(dds., name="treatment_AE_1_vs_Baseline")
summary(re.ae.b.1) 
```

For the other coefficients. 
```{r,message=FALSE}
res.ae.b.4 <- results(dds., name = "treatment_AE_4_vs_Baseline")
res.re.b.1 <- results(dds., name = "treatment_RE_1_vs_Baseline")
res.re.b.4 <- results(dds., name = "treatment_RE_4_vs_Baseline")
```

## Log fold change shrinkage for visualization and ranking
`lfcShrink` performs shrinkage of effect size, which is useful for visualization and ranking of genes. Again, using `summary` to obtain basic results. 
```{r,echo=TRUE}
#ae 1hr
resLFC_ae.1hr <- lfcShrink(dds.,
                           coef=2,
                           res = re.ae.b.1) 
summary(resLFC_ae.1hr)
```

Now, for the other coefficients. 
```{r,message=FALSE}
#ae_4hr
resLFC_ae.b.4 <- lfcShrink(dds.,
                           coef=3,
                           res = res.ae.b.4) 
#re 1 hr
resLFC_re.b.1 <- lfcShrink(dds.,
                           coef=4,
                           res = res.re.b.1)
#re 4 hr
resLFC_re.1.4 <- lfcShrink(dds.,
                           coef=5,
                           res = res.re.b.4)
``` 

# Exploring results
## MA-plot
`Plot MA` shows the log2 fold changes due to a given variable over the mean of normalized counts for all samples in the DESeqDataSet. Points are coloured red if the adjusted p value is lower than 0.1. Here, we plot our results obtained from `results` ("Normal"), followed by the shrunken results obtained from `lfcShrink` ("apeglm). 
```{r,message=FALSE}
xlim <- c(1,1e5); ylim <- c(-11,11) #Define limits of plot 

#ae.1hr
plotMA(re.ae.b.1, ylim=ylim,main="Normal") 
plotMA(resLFC_ae.1hr, xlim=xlim,ylim=ylim,main="apeglm")

#ae.4hr
plotMA(res.ae.b.4, ylim=ylim,main="Normal") #not run 
plotMA(resLFC_ae.b.4,xlim=xlim,ylim=ylim,main="apeglm")
#re.1hr
plotMA(res.re.b.1, ylim=ylim,main="Normal")
plotMA(resLFC_re.b.1,xlim=xlim,ylim=ylim,main="apeglm")

#re.4hr
plotMA(res.re.b.4, ylim=ylim,main="Normal") 
plotMA(resLFC_re.1.4,xlim=xlim,ylim=ylim,main="apeglm")
```

## Filtering results
We create a function to filter our results based on specific criteria. Here, our thresholds are adjusted *P*-values less than 0.05 and absolute log2FoldChange values of 0.58. Our results functions provides us with a data frame of filtered results. 
```{r,message=FALSE}
my_results = function(DESeqResults_class){
  df = as.data.frame(DESeqResults_class)
  filter_padj = df[df$padj < 0.05,]
  filter_logFC = filter_padj[abs(filter_padj$log2FoldChange) > 0.58,]
  clean_df = subset(filter_logFC, !is.na(baseMean))
  r2c = rownames_to_column(clean_df, var = 'ensgene')
  return(r2c)
}
```

## Organizing gene annotations
We create a function to annotate our results. 
We annotate our genes according to the GRCh37.p13 assembly (same assembly used by Dickinson et al., 2018). 
```{r,message=FALSE}
my_mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL",
                    host="grch37.ensembl.org",
                    path="/biomart/martservice",
                    dataset="hsapiens_gene_ensembl")

my_annotations_dickinson = function(x3){
  y = getBM(attributes = c('ensembl_gene_id',
                           'external_gene_name',
                           'gene_biotype',
                           'chromosome_name',
                           'start_position',
                           'end_position',
                           'strand',
                           'description'),
            filters = c('ensembl_gene_id'),
            values = x3$ensgene,
            mart = my_mart)
}
```

## Applying our new functions 
We apply our new functions (results and annotations) to our results for each coefficient. 
```{r,message=FALSE}
#ae 1hr
resLFC_ae.1hr_df <- as.data.frame(resLFC_ae.1hr)
my_ae.1hr_df <- my_results(resLFC_ae.1hr_df)
my_ae.1hr_df_anno <- my_annotations_dickinson(my_ae.1hr_df)
final.ae_1hr <- left_join(my_ae.1hr_df,my_ae.1hr_df_anno,
                          by=c('ensgene'='ensembl_gene_id'))
```

We see our final results for the first coefficient below (Baseline vs. 1 hour post-AE). 
```{r,message=FALSE}
final.ae_1hr 
```

```{r,message=FALSE}
#ae_4hr
ae_1.4_my.r_df <- my_results(resLFC_ae.b.4)
ae_1.4_my.r_df_anno <- my_annotations_dickinson(ae_1.4_my.r_df)
final.ae_4hr <- left_join(ae_1.4_my.r_df,ae_1.4_my.r_df_anno,
                          by=c('ensgene'='ensembl_gene_id'))

#re 1 hr
re_b.1.r_df <- my_results(resLFC_re.b.1)
re_b.1.r_df_anno <- my_annotations_dickinson(re_b.1.r_df)
final.re_1hr <- left_join(re_b.1.r_df,re_b.1.r_df_anno,
                          by=c('ensgene'='ensembl_gene_id'))

#re 4 hr
re_1.4.r_df <- my_results(resLFC_re.1.4)
re_1.4.r_df_anno <- my_annotations_dickinson(re_1.4.r_df)
final.re_4hr <- left_join(re_1.4.r_df,re_1.4.r_df_anno,
                          by=c('ensgene'='ensembl_gene_id'))
```

# Exporting results into csv files 
Finally, we export our results into individual csv files, where lncRNAs are filtered from other genes and transcripts. 
```{r,message=FALSE, eval=FALSE}
write_csv(final.ae_1hr, file = "/Users/kassiaregnier/Documents/regnier_thesis_results/AE_b.vs.1.csv",
          append = FALSE) 
write_csv(final.ae_4hr, file = "/Users/kassiaregnier/Documents/regnier_thesis_results/AE_b.vs.4.csv",
          append = FALSE) 
write_csv(final.re_1hr, file = "/Users/kassiaregnier/Documents/regnier_thesis_results/RE_b.vs.1.csv",
          append = FALSE)
write_csv(final.re_4hr, file = "/Users/kassiaregnier/Documents/regnier_thesis_results/RE_b.vs.4.csv",
          append = FALSE)
```
